---
title: "Flow Analysis COMPASS Tutorial"
author: "Jolie Phan"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Directory structure
To directly use the code in this tutorial, you should create an RStudio project and set up your project directory as follows.

The `data/` folder should contain all `.xml` and `.fcs` files and the `out/` folder will contain the GatingSet and all COMPASS outputs.

```{r dir_structure, echo=FALSE, message=FALSE}
# You don't need to run this! See README or html for figure.
library(DiagrammeR)
grViz(diagram = "digraph flowchart {
  node [fontname = arial, shape = retangle, fontsize=8, height=0.1]
  dir1 [label = '@@1']
  dir2 [label = '@@2']
  dir3 [label = '@@3']
  dir4 [label = '@@4']
  dir5 [label = '@@5']
  dir6 [label = '@@6']
  dir7 [label = '@@7']
  dir1 -> dir2,dir3;
  dir2 -> dir4,dir5;
  dir3 -> dir6,dir7;
  }
  
  [1]: 'Rproject/'
  [2]: 'data/'
  [3]: 'out/'
  [4]: '.xml'
  [5]: '.fcs files'
  [6]: 'GatingSet/'
  [7]: 'COMPASSResult/'
  ")
```

You can achieve this directory structure by running the following command lines:

```{r}
if(!dir.exists(here::here("data"))) {
  cat(sprintf("Creating folder %s\n", here::here("data")))
  dir.create(here::here("data"), recursive = T)
}
if(!dir.exists(here::here("out"))) {
  cat(sprintf("Creating folder %s\n", here::here("out")))
  dir.create(here::here("out"), recursive = T)
}
if(!dir.exists(here::here("out/GatingSet"))) {
  cat(sprintf("Creating folder %s\n", here::here("out/GatingSet")))
  dir.create(here::here("out/GatingSet"), recursive = T)
}
if(!dir.exists(here::here("out/COMPASSResult"))) {
  cat(sprintf("Creating folder %s\n", here::here("out/COMPASSResult")))
  dir.create(here::here("out/COMPASSResult"), recursive = T)
}
```

# Load libraries

```{r}
library(here)
library(CytoML)
library(flowCore)
library(flowWorkspace)
library(COMPASS)
```

# Load data
Download the .xml file and folder containing the associated .fcs files (see the README for updated directions). Then, drag them into the "data" folder of the project directory.

This data has been gated in FlowJo v9 and will be parsed using flowWorkspace.

**FYI:** *When using FlowJo v9, the FlowJo workspace must be exported as an .xml file to create a flowjo_workspace object with the function open_flowjo_xml(). However, when using FlowJo v10, the FlowJo workspace can be loaded directly as a .wsp file using the same function open_flow_xml().*

```{r}
# Location of XML file
xml_path <- here::here("data/20200605_COVID_ICS-B3-trunc.xml")

# Location of .fcs files
fcs_subfolder <- here::here("data/20200605_COVID_ICS-B3-FCS-trunc/")
```

Create a flowjo_workspace object with the function open_flowjo_xml().

```{r}
ws <- open_flowjo_xml(xml_path)
```

# Create a GatingSet

## Set-up

A GatingSet holds a set of GatingHierarchy objects, representing a set of samples 
and the gating scheme associated with each.

Look at the workspace metadata to choose which keywords to extract into the 
GatingSet. The flowjo_to_gatingset() function parses a flowJo Workspace to 
generate a GatingSet object.

```{r}
# Look at all of the keywords
names(fj_ws_get_keywords(ws, 117)) 
# Choose which keywords to keep
keywords2import <- c("EXPERIMENT NAME",
                       "$DATE",
                       "SAMPLE ID",
                       "PATIENT ID",
                       "STIM",
                       "WELL ID",
                       "PLATE NAME") 
sampleGroup <- "Samples"

gs <- flowjo_to_gatingset(ws,                                    
                          name = sampleGroup, 
                          keywords = keywords2import,
                          path = fcs_subfolder, 
                          extend_val = -10000)
```

## QC

Make sure that the gating trees are consistent for all samples.

```{r}
pop_lists <- lapply(gs, gh_get_pop_paths)
unique(pop_lists)
```

Remove channels from flow data that are not used by gates.

```{r}
gs <- gs_remove_redundant_channels(gs) # drop SSC-H
```

Add names to all channels or change their names.

```{r}
dput(unname(pData(parameters(gh_pop_get_data(gs[[1]])))[,2]))
markernames <- c("Time", "FSC-A", "FSC-H", "SSC-A", "CD8b", "TNFa", "CD107a",
                 "CD154", "CD3", "IL2", "CD4", "IL17a", "IL4_5_13", "CD14_19",
                 "CCR7", "CD38", "LD", "IFNg", "CD45RA", "HLADR")
names(markernames) <- pData(parameters(gh_pop_get_data(gs[[1]])))[,1]
markernames(gs) <- markernames
pData(parameters(gh_pop_get_data(gs[[1]])))[,c(1,2)]
```

In this tutorial, we will only run COMPASS on CD4+ T cells, but what if we wanted to also run COMPASS on CD8+ T cells? Currently, most of the gates are missing from the "8+" node. Let's grab the missing gates from the "NOT4+" node and add them under the "8+" node.

```{r}
gates_to_copy <- c("/Time/LD-3+/1419-3+/S/Lymph/NOT4+/107a",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/154", 
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/IL2",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/IL17",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/IL4513",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/TNF",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/CCR7+",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/CD45RA+")

for(path in gates_to_copy) {
  gs_pop_add(gs, lapply(gs, gh_pop_get_gate, y=path),
             parent = "/Time/LD-3+/1419-3+/S/Lymph/8+")
}
recompute(gs, "/Time/LD-3+/1419-3+/S/Lymph/8+")
```

Next, let's get rid of the "NOT4+" node and all of its descendants.

```{r}
gs_pop_remove(gs, "/Time/LD-3+/1419-3+/S/Lymph/NOT4+")
```

Plot the gating tree.

```{r, fig.width = 6, fig.height = 4}
plot(gs, fontsize=15, bool=T)
```



## Save GatingSet

**Note:** this can take a while. 

```{r}
save_gs(gs, here::here("out/GatingSet"))
```

# Create a COMPASSContainer

A COMPASSContainer is the data structure used to hold data from an ICS experiment.
The input for this code is a GatingSet or a GatingSetList. Counts, metadata, and 
single cell data are extracted and fed into the COMPASSContainer constructor.

```{r}
# Set the seed
set.seed(123)

# A regular expression to match a single node in the gating tree
parent_node <- "4+"

# A character identifying the subject id column in the GatingSet metadata
id <- "SAMPLE ID"

markernames(gs) 

# markermap contains the output of markernames(gs)
markermap <- list("IL2", "IL4_5_13", "IFNg", "TNFa", "IL17a", "CD154", "CD107a")

# Assign names to the list of markers based on the gate names
names(markermap) <- paste0(parent_node, "/", c("IL2", "IL4513", "IFNG",
                                               "TNF", "IL17", "154", 
                                               "107a"))
```

Construct the COMPASSContainer. If the number of parent cells is less than countFilterThreshold, we drop that file (default is 5000 cells).

```{r}
CC <- COMPASSContainerFromGatingSet(gs,
                                    node = parent_node,
                                    individual_id = id,
                                    mp = markermap,
                                    countFilterThreshold = 5000)
```

Look at some basic info about our COMPASSContainer.

```{r}
CC
```

# Run COMPASS

Fit the COMPASS model using the COMPASSContainer. To fit the COMPASS model, we 
need to specify how to identify the samples that are our treatment condition and 
our control condition. Here, we will run COMPASS on the samples stimmed by spike
1 with DMSO as our negative control. For now, let's just do 100 iterations for 
speed.

```{r}
fit <- COMPASS(CC,
               treatment = STIM == "Spike 1",
               control = STIM == "DMSO",
               iterations = 100)
```

Save the COMPASS run output.

```{r}
saveRDS(fit, file.path(here::here("out/COMPASSResult"), "COMPASSResult.rds"))
```

Save the Functionality and Polyfunctionality Scores.

```{r}
FS <- FunctionalityScore(fit)
FS_df <- data.frame(tmp = names(FS), FS = FS)
colnames(FS_df) <- c("SAMPLE ID", "FS")

PFS <- PolyfunctionalityScore(fit)
PFS_df <- data.frame(tmp = names(PFS), PFS = PFS)
colnames(PFS_df) <- c("SAMPLE ID", "PFS")

FS_PFS_df <- merge(FS_df, PFS_df, by = "SAMPLE ID")
write.table(FS_PFS_df,
            file = file.path(here::here("out/COMPASSResult"), "FS_PFS.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
FS_PFS_df
```

# Visualize

Plot a heatmap of the mean probability of response.

```{r, fig.width = 6, fig.height = 4}
plot(fit, show_rownames = TRUE)
```

